---
title: Dataset Basics
author: <a href="http://stanford.edu/~ejdemyr/"> Simon Ejdemyr </a> 
date: December, 2015
summary: This tutorial introduces datasets --- "data frames" in R. If you've
 completed the <a href="../vectors">tutorial on vectors</a>, you'll
 soon see that datasets can be thought of as a collection of vectors
 stored as columns in a dataset. We'll talk about how to create
 datasets and how to read them from file. We'll also talk more
 conceptually about how datasets should be structured.
output:
  html_document:
     template: ../template/r-tutorial-template-v2.html
     mathjax: null
---



Creating datasets
==================

Let's start by learning how to create a dataset in R. This turns
out to be very simple --- just combine vectors using the `data.frame()`
command. 


```{r}
# Create three vectors 
name <- c("al", "bea", "carol")
age <- c(6, 7, 4)
hair <- c("brown", "green", "blond")

# Create data frame 
children <- data.frame(name, age, hair)
children

# Creating a data frame can also be done without first saving vectors 
children <- data.frame(
    name = c("al", "bea", "carol"),
    age = c(6, 7, 4),
    hair = c("brown", "green", "blond")
    )
children
```

We created a dataset called `children`, which has 3 rows and 3
columns. We used two approaches that differ in whether they first save
vectors to R's memory. 


Dataset structure
====================

More important than learning the mechanics of creating a dataset in
R is to understand their general structure: 

1. Each column should consist of a vector that gives some fact about
the world (e.g., age in years). We usually refer to these columns
as *variables*. 
2. At least one column should identify who or what the information in
the data is about.  Such a variable is called an "id" variable or
"key". In the `children` dataset above this variable is `name`. The
remaining variables have the facts or measurements that we care
about. For example, we gather from the dataset that Al is 6 years old
(one fact) and that Al has brown hair (a second fact).

To better understand the proper structure of datasets, let's create a
second data frame. Suppose here that `gdp_pc` is a measure of a
country's GDP per capita in a given year. (Use `?expand.grid` and
`?runif` to learn more about these functions, though that is not a
priority right now.)

```{r}
countries <- data.frame(
    expand.grid(country = c("USA", "China", "Sudan"), year = 1994:1996),
    gdp_pc = round(runif(9, 1000, 20000), 0)
    )
countries
```

This time around our dataset has two id variables: `country` and
`year`. Why two and not one? One way to think about it is that
`country` by itself wouldn't be sufficient to uniquely identify a row,
because there are three rows for each country (and likewise with
`year`). Combined, however, `country` and `year` uniquely identify
each row. In other words, GDP per capita (the only fact or measurement
in this dataset) describes a given country in a given year.

We can say that the **unit of analysis** in the dataset `countries` is
country-year. This means that two id variables (country and
year) are required to uniquely identify each row. In the `children`
dataset above the unit of analysis is "child" or "person".


Basic commands
=================

Here are some commands that are useful for getting to know your data
and for understanding dataset structures in general.

Dimensions
---------------------

The first is `dim()`, which gives the dimensions of a data frame. The
number of rows are listed first, columns second.

```{r}
dim(countries)
```

Use `nrow()` and `ncol()` to to get the number of rows or columns
separately.  These commands are useful for
[code generalization](../vectors/#code-generalization).

```{r}
nrow(countries)
ncol(countries)
```

Snapshots
-------------

Use `head()` and `tail()` to look at the first and last few rows of
a dataset, respectively. This is more useful when we have datasets
with many observations. 

```{r}
head(countries)
tail(countries)
```

Other useful commands to get to know variables better include `summary()`,
`table()`, and `prop.table()`. 

```{r}
# Get some summary information about each variable
summary(countries)

# Number of observations by country 
table(countries$country)

# Proportion of observations by country 
prop.table(table(countries$country))
```

Accessing specific rows and columns
-------------------

Like with vectors, brackets (`[]`) can be used to access data in
datasets. But unlike with vectors, we need to input *two* arguments
--- separated by a comma --- into the brackets. The first argument
always applies to rows while the second applies to columns. 

```{r}
countries <- data.frame(
    expand.grid(country = c("USA", "China", "Sudan"), year = 1994:1996),
    gdp_pc = round(runif(9, 1000, 20000), 0)
    )
countries

# Access row 2, col 3
countries[2, 3]

# Access entire row 5
countries[5, ] #note: leaving second argument blank

# Access entire column 3
countries[, 3] #note: leaving first argument blank
```

In general, though, accessing rows and columns by index is bad for
[code generalization](../vectors/#code-generalization).  It
particularly causes problems when you add or delete rows/columns,
because then the indexing will change (e.g., column 3 representing GDP
per capita may now be in column 4).

For this reason, it's better to access columns using column
names. 

```{r}
# Access a column using column/variable name (two equivalent approaches)
countries$year
countries[, "year"]
```
Note that when we're accessing a column this way, it's just a vector
and all the things we've learned about [vectors](../vectors) apply.
For example:

```{r}
# Get mean gdp per cap
mean(countries$gdp_pc)
```

To access rows, it's best to use a logical statement, which is covered
in more detail in a separate tutorial on
[modifying data](../modifying-data).  But just to
give an example, here's how we can access a row using bracket notation
and a logical statement:

```{r}
countries[countries$year == 1995 & countries$country == "USA", ]
```



Reading data 
=========================


Note: In this section we'll be working with a dataset called
`world-small.csv`, which you can download [here](../data/world-small.csv).

So far we've created datasets ourselves. Oftentimes, however, we'll
want to read a dataset into R from file. Datasets come in many formats
--- e.g., .csv, .txt, .dta, and .RData. R can read most data formats
as is, but sometimes it may be necessary to manually reformat some
elements in the file or even to reconvert the whole file to a
different format (e.g., using Stat/Transfer). For now, we'll assume
that the file is in a readable format.

To read a file you need to 

1. Specify where the file is located on your computer. This is referred to 
as setting your working directory. 
2. Execute a command that will read the file from your working directory. 


Setting the working directory 
----------------------

You can set your working directory manually. In RStudio, go to
Session --> Set Working Directory --> Choose Directory...
and find the folder in which your file is located.

While this works, you should also set the working directory using
code. Use `setwd(path-to-dir)` where `path-to-dir` is the the path to
the folder in which the file is located. How can you find this path?
Here are instructions for
[Windows](https://technet.microsoft.com/en-us/magazine/ff678296.aspx)
and [mac](http://josharcher.uk/code/find-path-to-folder-on-mac/).
If you're still not sure how to do this, take a look at
[this tutorial](https://sites.google.com/site/manabusakamoto/home/r-tutorials/r-tutorial-4).

To check that your working directory includes the file you want
to read, use `dir()` without anything in the parentheses. This function
outputs all the files in your working directory into the R console.
So, if you want to read the `world-small.csv` file that you downloaded above,
you should see this file listed when you execute `dir()`. 


Reading the file 
----------------------

Now that we've told R where to look for our file, it's time to read
it. Different commands are used to read different types of files. This
is the syntax used for reading a .csv file:

```{r, echo=FALSE}
world <- read.csv("../data/world-small.csv")
```

```{r, eval=FALSE}
world <- read.csv("world-small.csv") 
```

I'm reading the file from the working directory and assigning it
to the object `world`, which becomes of class `data.frame`. 

```{r}
class(world)
```

Let's check if the file was read correctly, using `dim()`
(returns the dimensions), `head()` (returns the top six rows),
and `summary()` (returns summary information about each variable):

```{r}
dim(world) #the number of rows and columns 
head(world) #the first few rows of the dataset
summary(world) #a summary of the variables in the dataset
```

Everything looks as we would have hoped.


Exercises
============

1. Read the `world-small.csv` data into R and store it in an object
called `world`. (Set your working directory using code first.) 

2. (Conceptual) What is the unit of analysis in the dataset? What's the name
of the dataset's id variable?

3. How many observations does `world` have? How many variables? Use an R
command to find out.

4. Use brackets and a logical statement to inspect all the values for
   Nigeria and United States. That is, your code should return two
   entire rows of the dataset. 

5. Use R to return China's Polity IV score. As in question 4, use a logical
statement and brackets, but don't return the entire row. Rather, return a single
value with the Polity IV score.

6. What is the lowest GDP per capita in the dataset? (Use R to return only the value.)

7. What country has the lowest GDP per capita? (Your code should
return the country name and be general enough so that if the observations
in the dataset --- or their order --- change, your code should still return the
country with the lowest GDP per capita.)






