---
title: Introduction
author: <a href="http://stanford.edu/~ejdemyr/"> Simon Ejdemyr </a> 
date: December, 2015
summary: In this first in a number of tutorials, we'll cover the very basics of
 R. If you've programmed before you can skip much of this. But
 regardless of your background, I hope you'll find this and subsequent
 tutorials useful for learning R's many tools for graphing,
 statistical analysis, and data collection and management --- or what we
 collectively might call "data science." 
output:
  html_document:
     template: ../template/r-tutorial-template-v2.html
     mathjax: null
---


	

Running R
=================

You can run R using a number of text editors or "integrated
development environments" (IDEs). Most people prefer some other
application than R's native environment, which provides only limited
functionality in terms of syntax highlighting, auto-completion, and
debugging. Alternatives include [RStudio](https://www.rstudio.com/)
and
[Emacs](https://www.gnu.org/software/emacs/)/[ESS](http://ess.r-project.org/). I
very much prefer the latter, but if you've never programmed before I
would go with RStudio (installation instructions
[here](https://www.rstudio.com/products/rstudio/download/); note that
you need both R and RStudio).

All IDEs include a console and a text editor. The console is where you'll
see the results (or *output*) of commands executed from the editor. You can
type commands directly into the console, but this is generally not a good
strategy. This is because the whole purpose of writing code is to make
it reproducible. Typing commands in the text editor will let you
come back to them later as long as you save them (using extension .R).


Your first script
======================

Fire up RStudio (or whatever IDE you use) and type the following two
lines into the text editor. Then execute them, which can be done using a
[keyboard shortcut](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)
--- <span class="fw">Ctrl+Enter</span> (Windows) or <span
class="fw">Command+Enter</span> (mac) in RStudio.

```{r}
hello_world <- "hello world"
hello_world 
```

In case it's not obvious, R output is displayed in
<span style="color: #969696;"> gray</span>, above which is code that you
should type into your text editor.

Now execute the following code (but type it into your text editor first): 
```{r eval=FALSE}
# Create two objects and perform simple arithmetics
a <- 5
b <- 7
a - b
a + b + 10

# Create a third object 
w <- a + b
w

# Logical tests 
w - b == a
w - b > a
b <- 1             #overwrite b 
x <- w - b > a
x
```

Does the output make sense to you? (I'm not displaying it here so that
you can try to make sense of it yourself first.)

Here are some things to pay attention to from our first few lines of code:

1. **Comments**. Anything after `#` is a comment and won't be evaluated by R.

2. **The assignment operator**. R uses `<-` (the assignment
operator) to create objects. A single equal sign (`=`) also works, but
the norm is to use `<-` in R. 

3. **Objects**. We used the assignment operator to create *objects*
--- for example, `hello_world` and `a`. The names of these objects are
completely arbitrary. Unless they are created within a function, they
are stored in R's memory with the value you assigned to them (e.g.,
`"hello world"`, `5`) until the end of your R session.

4. **Class.** R automatically assigns a class to each object
you create. Three classes were used above: numeric,
character, and logical. Use `class()` to see what type an object is
(e.g., `class(hello_world)`, `class(a)`, and `class(x)`). Why is this
important? For one, you may think a certain object has one class when
it has another. This will cause issues. For example, this throws an error: 

<div class="indent">
```{r eval=FALSE}
a <- 5
b <- "7"
a + b 
```
Not knowing that `b` is a character object (which often is much
less obvious than here) would be frustrating. Other classes include
<span class="fw">factor</span>, <span class="fw">matrix</span>,
<span class="fw">data.frame</span> and many more that we'll get to. 
</div>

5. **Overwriting objects**. You can overwrite objects without R
complaining (e.g., `b <- 1` above). This can be confusing if you don't
keep track of which objects you have overwritten, but in general this
ability is a good thing. R stores every object you create in
memory, and overwriting an object saves memory to the extent that R
doesn't need to hold an additional object, which may be consequential
when you work with large datasets.


Packages 
========================== 

The true power of R is that it's open-source. As such, anyone can
extend its core functionality through packages. This often results in
remarkable improvements to how we can approach complex data tasks. In
subsequent tutorials, I will make particular use of a set of R packages 
developed by [Hadley Wickham](http://had.co.nz/). 

To use a package you must: 

1. **Install it.** You only need to do this once. 
2. **Load it.** You need to re-load packages every time you open R. 


To install packages `plyr`, `dplyr`, and `tidyr`, which we'll use a lot,
run

```{r eval=FALSE}
install.packages(c("plyr", "dplyr", "tidyr"), dep = T)
```

And to load these packages, use: 

```{r, comment = "", eval=FALSE}
require(plyr)
require(dplyr)
require(tidyr)
```

Or you can load many packages more compactly like so: 

```{r, comment = "", eval=FALSE}
pkgs <- c("plyr", "dplyr", "tidyr")
sapply(pkgs, require, character.only = T)
```

<div style="margin-top: 15px"> </div>

Style
======================

Writing good code requires a lot of practice. Good code executes
quickly and is non-repetitive (e.g., uses functions for tasks that
need to be executed several times). But most importantly for you right
now, good code is easy to understand --- both for yourself if you were
to come back to it and for someone else reading it.

To that end, comments are extremely useful. Comments can delineate
different blocks of code and, more importantly, clarify what a command
or set of commands is doing. Especially use comments if you're doing
something complex --- save them for routine tasks that most R
programmers will understand.

There are some other style guidelines you should attempt to follow.
Please take a look at
[Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml).
In particular, try to observe the spacing and line lengths limits
outlined in that document. (Some other rules seem less important to
me: I personally like underscores for object and function names.)




Trouble shooting 
========================== 

Many, many times when coding you'll have an idea of what you want to
do but won't know how to do it in R. This happens even for
experienced coders. With the right strategies, you'll be able to solve
a majority of issues you run into yourself. Not having to ask someone
else every time you run into a problem will save you a lot of time.

When you get stuck, google is your friend. For example, if you want
to find the mean of a variable, try googling "how to find mean in R"
and there will be tons of explanations of how to do this.

R also has a help feature that can be called using the following
syntax: `?commandname`, where `commandname` is the name of the command
that you need help with. For example, `?mean` will bring up a help
dialog box with information about how to use R's `mean()` command.


