---
title: "R Tutorial 4"
output:
  html_document:
    toc: true
    theme: united
    number_sections: true
---

* * * * * * * * * * * * * * * * * * * * * * * * 

Tables and Graphs for Professional Reports
==========================

Many of you will have to produce professional reports and
presentations, for which producing nice tables and graphs is
important. The best way to automate this process is to use
[LaTeX](http://www.latex-project.org/) and
[Beamer](http://en.wikibooks.org/wiki/LaTeX/Presentations) rather than
Microsoft Word and Powerpoint. Alternatively, check out the RStudio
team's slick [R Markdown package](http://rmarkdown.rstudio.com/),
which makes producing beautiful reports really simple.

However, here are some options if you still want to use Word. 

* * * * * * * * * * * * * * * * * * * * * * * * 

Tables
----------

You can produce tables quickly without having to copy and paste
every number from R to Word. 

To do so: 

1. Create a table or data.frame in R. 
2. Write this table to a comma-separated .txt file using `write.table()`. 
3. Copy and paste the text in the .txt file into Word. 
4. In Word, 
    a. select the text you just pasted from the .txt file
    b. go to Table $\rightarrow$ Convert $\rightarrow$ Convert Text to Table...
    c. make sure "Commas" is selected under "Separate text at", click OK 

The text should now end up in a table that you can format in Word.  

### Example 1 ###
Here's an example of the first two steps using a [dataset on
U.S. states](states.csv) (codebook [here](states_codebook.csv)).

```{r}
# Read data
states <- read.csv("states.csv")          

# (1) Create a table of Bush support by U.S. region in 2000 (South versus Non-South): 
t <- with(states, table(south, gb_win00))
t <- prop.table(t, margin = 1)
t                                         #large Majority of Southerners voted for Bush:    

# (2) Write this table to a comma separated .txt file:    
write.table(t, file = "bush_south.txt", sep = ",", quote = FALSE)
```
The .txt file will end up in your working directory. Now follow steps 3 and 4
to create a table in Word. 

### Example 2 ###
Here's another example that again uses the [states.csv](states.csv) dataset. Say we 
wanted to create a table with summary statistics for five of the variables
in this dataset: 

```{r}
# Keep 5 variables in states dataset
states_sub <- select(states, blkpct, attend_pct, bush00, obama08, womleg)

# Find summary statistics for each variable
library(plyr)                              #to access colwise function
means <- colwise(mean)(states_sub)
stdev <- colwise(sd)(states_sub)
mins <- colwise(min)(states_sub)
maxs <- colwise(max)(states_sub)

# Create df with summary statistics, putting variables in rows using transpose function t()
df <- data.frame(t(means),
                 t(stdev),
                 t(mins),
                 t(maxs))

# Clean column and row names
names(df) <- c("Mean", "SD", "Min", "Max")
row.names(df) <- c("Black (%)", "Attend Church (%)", "Bush -00 (%)",
                   "Obama -08 (%)", "Women in Legislature (%)")

# Restrict number of decimal points to 1
df <- round(df, 1)
df

# Write data frame to .txt file
write.table(df, file = "sumstats.txt", sep = ",", quote = FALSE)

```

### Exercises ###

1. Create a table of summary statistics in Word for `vep04_turnout`,
`vep08_turnout`, `unemploy`, `urban`, and `hispanic`. The table should
include the number of observations (*n*), mean, median, 10th
percentile, and 90th percentile of each of the variables. Put the
variables in the rows of the table and the summary statistics in the
columns, like we did in the example above. Format your table in Word
to make it look similar to [this table](summary_stats_example.tif). 


* * * * * * * * * * * * * * * * * * * * * * * * 

Graphs
---------

In Tutorial 2, we covered graphing with the `ggplot` package. Let's talk about 
how to ensure that the graphs you produce look good when you include them 
in your write-ups. 

### Saving images as .pdf ###

Saving images as .pdf is usually your best option. This format ensures that 
images don't pixelate. (And you can insert .pdfs into word like you do with 
other image file formats.) 

To save a .pdf, use the `pdf()` function before the image you want to save, and
include `dev.off()` after the image. 

Here's an example, again using the [states.csv](states.csv) dataset: 

```{r}
states <- read.csv("states.csv")

library(ggplot2)
p <- ggplot(states, aes(x = attend_pct, y = bush00)) +
       geom_point() +
       geom_text(aes(label = stateid, y = bush00 - 0.7), size = 3) +
       geom_smooth(method = "loess", se = F) +
       xlab("% in State Attending Religious Services") +
       ylab("% in State Voting for Bush in 2000")

# Save the image as a pdf:
pdf(file = "bush_religion.pdf", height = 6, width = 8)
p
dev.off()
```

### Arranging images in columns and rows ###

Arranging graphs into a matrix of rows and columns, like we did on problem set
2, can be very useful for presentational purposes. There are two ways to do 
this using `ggplot`: 

1. Create each graph separately and then arrange them using the function 
`grid.arrange()` in the `gridExtra` package. 
2. Use
[facet_wrap()](http://docs.ggplot2.org/0.9.3.1/facet_wrap.html), like
we did in [this example in Tutorial
2](http://stanford.edu/~ejdemyr/tutorial2.html#scatter-plots).  This
second approach is useful if we want to display the same relationship
across different groups or years.

Here's an example of the first approach: 

```{r fig.width=8, fig.height=8}
p1 <- ggplot(states, aes(x = bush00, y = bush04)) +
        geom_point() +
        geom_text(aes(label = stateid, y = bush04 - 0.7), size = 3) +
        geom_smooth(method = "loess", se = F) +
        xlab("% in State Voting for Bush in 2000") +
        ylab("% in State Voting for Bush in 2004")  

p2 <- ggplot(states, aes(x = bush04, y = obama08)) +
        geom_point() +
        geom_text(aes(label = stateid, y = obama08 - 0.7), size = 3) +
        geom_smooth(method = "loess", se = F) +
        xlab("% in State Voting for Bush in 2004") +
        ylab("% in State Voting for Obama in 2008")

p3 <- ggplot(states, aes(x = vep04_turnout, y = bush04)) +
        geom_point() +
        geom_text(aes(label = stateid, y = bush04 - 0.7), size = 3) +
        geom_smooth(method = "loess", se = F) +
        xlab("Turnout among Voting Eligible Population (2004)") +
        ylab("% in State Voting for Bush in 2004")

p4 <- ggplot(states, aes(x = vep08_turnout, y = obama08)) +
        geom_point() +
        geom_text(aes(label = stateid, y = obama08 - 0.7), size = 3) +
        geom_smooth(method = "loess", se = F) +
        xlab("Turnout among Voting Eligible Population (2008)") +
        ylab("% in State Voting for Obama in 2008")

library(gridExtra)
grid.arrange(p1, p2, p3, p4,     #specify the graphs to include
             ncol = 2)           #specify the number of columns we want   

```

Of course, you could save this graph using the `pdf()` function from above. 

### Exercises ###

1. Using `ggplot` and `gridExtra`, create four scatterplots of your choice 
(not the same as in the examples above) and arrange them into 2 rows and 
2 columns. 
2. Save this image using `pdf()` and 'dev.off()`, specifying an appropriate 
width and height, and insert this image into Word. 



* * * * * * * * * * * * * * * * * * * * * * * * 

Hypothesis Testing in R
==========================

For Problem Set 3, you will need to carry out one- and two-sample hypothesis
tests. Refer to the lecture notes for the theory behind these tests. What
follows is a brief discussion of how to implement these tests in R. 
Let's keep working with the [states.csv dataset](states.csv). 

* * * * * * * * * * * * * * * * * * * * * * * *

Chi-Squared Tests
---------------

Given a cross-tab, a chi-squared test essentially tests whether there is a 
"relation between the rows and columns", or whether there is statistical 
independence given the marginal distributions of the rows and columns. 

```{r}
states <- read.csv("states.csv")

with(states, table(gb_win00, states$gay_policy))

# Rearrange the order of the gay policy scale 
states$gay_policy <- factor(states$gay_policy,
                            levels = c("Most liberal", "Liberal",
                                       "Conservative", "Most conservative"))

with(states, table(gb_win00, states$gay_policy))
```

* * * * * * * * * * * * * * * * * * * * * * * *

**Class Exercise:** What would this distribution look like if the cell 
values approximately were proportional to the marginal distributions? 

* * * * * * * * * * * * * * * * * * * * * * * *

Let's do a chi-squared test on the actual distribution: 

```{r, warning = F}
t <- with(states, table(gb_win00, states$gay_policy))
chisq.test(t)
```

How do we interpret this output? 

* * * * * * * * * * * * * * * * * * * * * * * *

One-Sample t-Tests
---------------
In one-sample t-tests, we test whether an estimated mean can be statistically
distinguished from a posited "true" population mean $\mu_0$. Let's test whether per 
capita income---with an estimated mean of 31951 across
states---actually is 30000. So, the null hypothesis defines $\mu_0 =$ 30000. 
How weird would it be to see a value of 31951 given that 
$\mu_0$ actually is 30000? 

```{r}
mean(states$prcapinc)
t.test(states$prcapinc, mu = 30000)
```

How do we interpret this output? 

<!-- If the true mean---assumed to follow a t-distribution with 49 (*n* - 1)
degrees of freedom---actually was 30000, then we would expect to see a
value of 31951 only about 3 times in 1000, so we can be quite
confident that average per capita income is not 30000. -->
 

* * * * * * * * * * * * * * * * * * * * * * * *

Two-Sample t-Tests
---------------

In two-sample t-tests, we want to test whether two samples (or groups)
assumed to come from the same distribution have different means. For
example, say we wanted to test whether the percentage women in state
legislatures differ across Southern and non-Southern states. Before
we carry out this test, what is the null hypothesis? What is the alternative
hypothesis? 


The following carries out a Welch test, which doesn't assume that the two
groups have the same variances and uses Satterthaite-Welch adjustment
of the degrees of freedom (usually resulting in non-integer degrees of
freedom):

```{r} 
with(states, t.test(womleg ~ south))
```

How do we interpret this output? 


* * * * * * * * * * * * * * * * * * * * * * * * 
\

The code used in this tutorial is available [here](tutorial4.R). 
\
\








