---
title: Tables in R (And How to Export Them to Word)
author: <a href="http://stanford.edu/~ejdemyr/"> Simon Ejdemyr </a> 
date: October, 2015
summary: This tutorial explains how to create and export different
 types of tables in R. It was originally designed for undergraduate
 students at Stanford who tend to use Word rather than LaTeX.
output:
  html_document:
     template: ../template/r-tutorial-template-v2.html
     mathjax: null
---


Overview
==========================

How to export tables from R depends on what word processor you
use. This tutorial focuses on Word. If you use LaTeX, there are many
existing R packages and tutorials that will get you started, including
[xtable](https://cran.r-project.org/web/packages/xtable/xtable.pdf)
and
[stargazer](https://cran.r-project.org/web/packages/stargazer/stargazer.pdf).


To export tables to Word, follow these general steps: 
   
<div style="margin-left: 20px">    
   1. Create a table or data.frame in R. 
   2. Write this table to a comma-separated .txt file using `write.table()`. 
   3. Copy and paste the content of the .txt file into Word. 
   4. In Word, 
     a. select the text you just pasted from the .txt file
     b. go to Table $\rightarrow$ Convert $\rightarrow$ Convert Text to Table...
     c. make sure "Commas" is selected under "Separate text at", click OK 
</div>

You'll now have a basic table that you can format in Word. Below are
three examples of how to use this process to create crosstabs, tables
for summary statistics, and regression tables.

Data and Packages
--------------------
Before we get started, read in a 
[dataset on U.S. states](../data/states.csv) (codebook
[here](../data/states_codebook.csv)) into R: 

```{r, echo=FALSE}
states <- read.csv("../data/states.csv")
```

```{r, eval=FALSE}
states <- read.csv("states.csv")
```

Also install and load packages `dplyr`, `tidyr`, and `broom`: 

```{r, eval=FALSE}
pkgs <- c("dplyr", "tidyr", "broom")
install.packages(pkgs) #install 
sapply(pkgs, require, character.only = T) #load 
```

```{r, echo=FALSE}
pkgs <- c("dplyr", "tidyr", "broom")
sapply(pkgs, require, character.only = T) #load 
```


Crosstabs
====================

Create a table showing the proportion of states that supported Bush in
2000, by region (South versus Non-South):

```{r}
# Create table
t <- with(states, table(south, gb_win00))
t <- prop.table(t, margin = 1)

t #large majority of southern states supported Bush in 2000:   

# Write this table to a comma separated .txt file:    
write.table(t, file = "bush_south.txt", sep = ",", quote = FALSE, row.names = F)
```
The .txt file will end up in your working directory. Now follow steps 3 and 4
in the Overview section above to create the crosstab in Word. 

Summary statistics 
=======================

Here's another example that again uses the [states.csv](states.csv) dataset. Say we 
wanted to create a table with summary statistics for five of the variables
in this dataset: 

```{r}
sumstat <- states %>%

    # Select and rename five variables 
    select(
        `Black (%)` = blkpct,
        `Attend church (%)` = attend_pct,
        `Supported Bush in 2000 (%)` = bush00,
        `Supported Obama in 2008 (%)` = obama08,
        `Women in State Legislature (%)` = womleg
        ) %>%

    # Find the mean, st. dev., min, and max for each variable 
    summarise_each(funs(mean, sd, min, max)) %>%

    # Move summary stats to columns
    gather(key, value, everything()) %>% 
    separate(key, into = c("variable", "stat"), sep = "_") %>%
    spread(stat, value) %>%

    # Set order of summary statistics 
    select(variable, mean, sd, min, max) %>%

    # Round all numeric variables to one decimal point
    mutate_each(funs(round(., 1)), -variable)

sumstat

# Write to .txt
write.table(sumstat, file = "sumstats.txt", sep = ",", quote = FALSE, row.names = F)

```

Again, the `sumstats.txt` file will end up in your working directory,
and you can use steps 3 and 4 from the Overview section above to
import this file into Word.


Exercise 
-------------

Create a table of summary statistics in Word for `vep04_turnout`,
`vep08_turnout`, `unemploy`, `urban`, and `hispanic`. The table should
include the number of observations (*n*), mean, median, 10th
percentile, and 90th percentile of each of the variables. Put the
variables in the rows of the table and the summary statistics in the
columns, like we did in the example above. Format your table in Word
to make it look similar to [this table](summary_stats_example.tif). 


Regression tables
=======================

Say we wanted to run three OLS models to predict state-level support
for Bush in 2000, where each model adds a predictor to the preceding
model. We can create a regression table with all three models like so: 

```{r}
m1 <- tidy(lm(bush00 ~ blkpct, states)) 
m2 <- tidy(lm(bush00 ~ blkpct + south, data = states))
m3 <- tidy(lm(bush00 ~ blkpct + south + womleg, data = states))
# Note that tidy() from the broom package is used to convert each model to a data frame

all_models <- rbind_list(
    m1 %>% mutate(model = 1),
    m2 %>% mutate(model = 2),
    m3 %>% mutate(model = 3))

all_models


# Now make this data frame look more like a regression table 
ols_table <- all_models %>%
    select(-statistic, -p.value) %>%
    mutate_each(funs(round(., 2)), -term) %>% 
    gather(key, value, estimate:std.error) %>%
    spread(model, value) 

ols_table

# Export
write.table(ols_table, file = "olstab.txt", sep = ",", quote = FALSE, row.names = F)

```

Again, follow steps 3 and 4 from the Overview section above to import
the content of the .txt file into Word. 





