---
title: Vectors
author: <a href="http://stanford.edu/~ejdemyr/"> Simon Ejdemyr </a> 
date: December, 2015
summary: This tutorial introduces a key object in R&#58; vectors. Vectors have one
 dimension that represents some concept or fact about the
 world. Examples include age, height, eye color, GDP per capita, and
 war initiation. The tutorial explains how to create different types of
 vectors, how to subset them, how to modify them, and how to summarize
 them.
related: <a href="../introduction">Introduction</a>
         <a href="../dataset-basics">Dataset basics</a> 
         <a href="../modifying-data">Modifying data</a>
output:
  html_document:
     template: ../template/r-tutorial-template-v2.html
     mathjax: null
---



Creating vectors
================

We'll create three types of vectors: numeric, character, and logical.
Here are examples of each type: 

```{r}
# Numeric vectors 
n1 <- 20
n2 <- c(20, 25, 60, 55)

# Character vectors
c1 <- "Blue"
c2 <- c("Red", "Green", "Purple")

# Logical vectors
l1 <- TRUE
l2 <- c(TRUE, FALSE, TRUE) 
```

Note that vectors can consist of one or many *elements*. Three
common ways to create vectors with more than one element is to use `c()`,
`seq()`, or `rep()`.

<span class="fw"> c() </span>
-------------------------

As illustrated above, one very common way to create vectors with more
than one element is to use `c()` ("concatenate"), which simply combines
whatever values you specify in the parentheses.

<span class="fw"> seq() </span>
-------------------------

`seq()` applies to numeric vectors only: 

```{r}
n1 <- seq(from = 0, to = 10, by = 2)   #using 'by'
n1
n2 <- seq(from = 0, to = 10, length.out = 5)   #using 'length.out'
n2
n3 <- seq(1, 2, 0.1)   #no argument names specified; automatically uses 'from', 'to', 'by'
n3
n4 <- 1:5    #shortcut for integer sequence; same as 'seq(1, 5, 1)'
n4
```

`seq()` by default takes three parameters: starting value, end
value, and a value that specifies how elements will be incremented
("by"), which can be substituted with "length.out". Integer sequences can
be created using a colon.

<div class="general-note">

<strong> General note about argument names: </strong> A function's
argument names need not be specified, as illustrated when we created
`n3`. If they are not specified, R uses arguments based on a default
order. One way to learn about this order is to use `?` (e.g.,
`?seq`). If you specify the argument names, the order doesn't matter.
Putting all this together, hopefully it's obvious why `seq(by = -1,
from = 10, to = 2)` is the same as `seq(10, 2, -1)`.

</div>


<span class="fw"> rep() </span>
------------------------------

Vectors can also be created using `rep()`. As the name implies, this
function is useful if you want to *repeat* an element or elements.

```{r}
rep(1, 5)
rep("blue", 3)
rep(TRUE, 4)
```

As should be obvious, the first parameter in the function specifies
the element to repeat, and the second the number of times to repeat it.


Using more than one function
----------------------------

Perhaps the most powerful use of these functions comes from combining them.
Here are a two examples:

```{r}
rep(c("blue", "red"), 3)
c(rep(seq(0, 6, 2), 2), 4:1) 
```

The second example is somewhat hard to follow, and is probably at the
limit of complexity in terms of how many functions we want to combine. Separating
a task into multiple lines of codes can help. 

```{r}
s <- rep(seq(0, 6, 2), 2)
c(s, 4:1)
```


Subsetting vectors
======================

Extracting a subset of elements from a vector is an extremely
important task, not least because it generalizes nicely to datasets
(which are at the heart of data science). This process --- whether
applied to a vector or a dataset --- is often referred to as "taking a
subset", "subsetting", or "filtering". If there is one skill you need
to master as quickly as possible, it's this.

In R, there are three ways to filter a vector: using a separate
logical vector, using indexing, and using names. I tend to use the
first method most, but all three are useful.

Subsetting with logicals 
---------------------

Let's jump right into an example. Say we have a character vector
with only two elements ("apple" and "banana"). Subsetting it to
"apple" could be done like so: 

```{r}
fruits <- c("apple", "banana")
fruits[c(TRUE, FALSE)]
```

Note the use of brackets, `[]` --- this is common when filtering.
Within these brackets is a vector with the same number of logical
elements as there are elements in the vector you want to
subset. Elements across the two vectors are matched by order: elements
that match with `TRUE` are kept while elements that match with `FALSE`
are dropped.

This process is extremely useful when combined with a *logical
operation*.  Please familiarize yourself with the logical operations
listed [here](http://www.statmethods.net/management/operators.html).
For example, using a logical operation we can filter a large vector of
oranges, apples and bananas:

```{r}
# Create a vector with 30 fruits 
fruits <- rep(c("orange", "apple", "banana"), 10)
fruits 

# Create a logical vector for dropping bananas
# Note: I'm creating the exact same logical vector three times (overriding it each time)
# This is for illustrative purposes; using one of these is sufficient
lv <- fruits == "orange" | fruits == "apple" 
lv <- fruits != "banana" 
lv <- fruits %in% c("orange", "apple")
lv

# Carry out the subset
fruits[lv]
```

We applied the same logic as above: We have a vector (`fruits`) that
we want to subset. We do so using a logical vector (`lv`), where
elements that match with `TRUE` are kept. The only difference here is
that we create the logical vector with a logical operation. The logical
operators (e.g., `!=`, `|`) used here are discussed in the link above,
with the exception of `%in%`. 

<div class="general-note">

<strong> General note about `%in%`: </strong> This operator is
extremely useful as an alternative for repeated "or" (`|`) statements.
For example, say you have a vector with 10 types of fruits and you
want to keep elements that are equal to "orange", "apple", "mango",
"mandarin", or "kiwi". You could accomplish this by creating a logical
vector like so: `lv <- fruits == "orange" | fruits == "apple" | fruits
== "mango" | fruits == "mandarin" | fruits == "kiwi"`.  <br> What a
nighmarishly long statement compared to the `%in%` option that
accomplishes the exact same thing: `lv <- fruits %in% c("orange",
"apple", "mango", "mandarin", "kiwi")`.

</div>

Of course, subsetting using logicals can also be done on numeric vectors.
Here are a few examples:

```{r}
# Create a numeric vector
numbers <- seq(0, 100, by = 10)
numbers

# Illustrate three different filters
numbers[numbers <= 50 & numbers != 30]
numbers[numbers == 0 | numbers == 100]
numbers[numbers > 100] #returns an empty vector
```

Note that I didn't create logical objects to carry out the subsets here,
as opposed to above where we explicitly defined `lv`. I find it more
compact and intuitive to take subsets without first creating a logical
vector.


Subsetting using indexing
---------------------

A different way to subset a vector is to specify the index or indeces
you want to keep, again using brackets. Here are a few examples:

```{r}
fruits <- c("apple", "banana")
fruits[1]
fruits <- rep(c("orange", "apple", "banana"), 10)
fruits[c(10, 20)]
fruits[seq(1, 30, by = 5)]
```

I sometimes use this when I want to inspect or modify an
element that I know occurs at a specific index in the vector, a more
manual approach than using logical statements. 

Subsetting using indexing can also be used in *random sampling*,
which has many important applications --- for example, in experiments
and when you want to test-run code on a representative subset of your
data. So, let's introduce the `sample()` function:

```{r}
# Draw 10 elements at random from 1 to 100 
sample(1:100, size = 10)
```

The function takes a vector of values (often successive integer
values) and an argument that specifies how many values to draw at
random from this vector.  We can use the resulting values as
indeces to subset another vector: 

```{r}
fruits <- rep(c("orange", "apple", "banana"), 10)
fruits[sample(1:30, size = 5)]
```

Here, we're drawing a random sample of five elements from the vector
`fruits`.  Why did I specify `1:30`? Well, `fruits` consists of 30
elements, so specifying something like `1:100` likely would have
resulted in sampled values outside the bounds of the vector (e.g.,
`fruits[35]` doesn't exist).  Specifying `1:30` gives every element in
`fruits` an equal chance of being included in the sample.

Subsetting using names
---------------------

Lastly, we can assign names to each element in a vector and take a
subset based on the names.

```{r}
age <- c(50, 55, 80)
names(age) <- c("mom", "dad", "grandpa")
age #note that values now have names 
age[c("dad", "grandpa")] #subset
```

That is, we have a vector representing the age of three family
members.  We assign names to each value, and then keep the values
associated with two of the family members.


Modifying vectors
=====================

The subsetting logic from above can be used to modify vectors. The
idea here is that instead of keeping elements that meet a logical
condition or occur at a specific index, we can change them. For example,
what if we had mis-entered grandpa's age above? We can fix it using indexing,
a logical statement, or naming. 

```{r}
# Recreate vector with age values from above
age <- c(50, 55, 80)
names(age) <- c("mom", "dad", "grandpa")

# Three ways of changing grandpa's age
# Note: you'd only need to use one of these
age[age == 80] <- 82 #using a logical statement
age[3] <- 82 #using indexing
age["grandpa"] <- 82 #using naming
age
```

A logical statement is most efficient when we need to change a lot
of elements.

```{r}
fruits <- rep(c("orange", "apple", "bamama"), 5) 
fruits #bamamas anyone? 

# Let's fix the misspelled element
fruits[fruits == "bamama"] <- "banana"
fruits
```

Vector arithmetics
--------------------

We can modify or create new numeric vectors using arithmetic
operations. Three common types of operations involve: 

1. A vector with more than one element and a vector with only one element.
2. Two vectors with the same number of elements. Elements are matched
based on index.
3. A vector modified by a function. 

In all cases, we can modify all elements of a vector or only a subset
of elements using the bracket notation we learned above. 


```{r}
numbers <- 1:10
numbers

# One value modifying all values in a vector 
numbers <- numbers / 10
numbers 

# One value modifying a subset of a vector 
numbers[numbers > 0.5] <- numbers[numbers > 0.5] * 100
numbers

# Two vectors with the same number of elements 
numbers1 <- 1:10
numbers2 <- 10:1
numbers3 <- numbers2 - numbers1
numbers3

# Replacing a subset of a vector using another vector
numbers <- 1:10
numbers[numbers > 5] <- 5:1
numbers

# Modify a vector (or a subset of a vector) using a function
numbers <- 1:10
sqrt(numbers) #square root
exp(numbers) #exponentiate
log(numbers[c(1, 5, 10)]) #natural log
```

Vector arithmetics can also be carried out in R on two multi-value
vectors with different number of elements. Such operations use
the [recycling rule](http://www.r-tutor.com/r-introduction/vector/vector-arithmetics).



Summarizing vectors 
===========================

We often want to get summary statistics from a vector --- that is,
learn something general about it by looking beyond its constituent
elements. If we have a vector in which each element represents a
person's height, we may want to know who the shortest or tallest
person is, what the median or mean height is, what the standard
deviation is, etc. Here are common summary facts for vectors:

```{r}
numbers <- sample(1:1000, 10)
numbers

class(numbers) #check the class
length(numbers) #number of elements 
max(numbers) #maximum value
min(numbers) #minimum value
sum(numbers) #sum of all values in the vector
mean(numbers) #mean
median(numbers) #median
var(numbers) #variance
sd(numbers) #standard deviation
quantile(numbers) #percentiles in intervals of .25 
quantile(numbers, probs = seq(0, 1, 0.1)) #percentiles in invervals of 0.1
summary(numbers) #function that contains many summary stats from above
```

If you forget one of these functions or if I haven't included one here
that you need, google almost surely has the answer for you. Also note
that some of the operations above --- most notably `class()` and
`length()` --- apply to non-numeric vectors.


Code generalization
-------------

We want our code to be as general as possible so that it can be
reapplied to a different coding task or if the data change.
Commands that summarize vectors can be useful to accomplish this. 

Remember above when we found a random sample of `fruits`?  Here is
more or less the code we used:

```{r}
fruits <- rep(c("orange", "apple", "banana"), 10)
length(fruits) 
random_sample <- fruits[sample(1:30, size = 5)]
random_sample
```

The third line, where we create `random_sample`, is not very
general. Why? In this case, `fruits` has 30 elements.  What if it
instead had 50 elements? *Then the third line would not give us a
random sample.* Or more precisely, this line would give us a random
sample of the 30 first elements of fruits --- the last 20 elements
would not have a chance of being included. We could modify the third
line to read `random_sample <- fruits[sample(1:50, size = 5)]`.  But
if we then modified `fruits` to have a different number of elements
again we'd end up with the same problem.

Here's the solution: find the number of elements of fruits using
`length()` and then input this as an argument in the `sample()`
function.

```{r}
fruits <- rep(c("orange", "apple", "banana"), 100)
n <- length(fruits) #store the result of length() in an object 
n

random_sample <- fruits[sample(1:n, size = 5)] #now use 'n' in the sample() function
random_sample

# Or we could have used length() directly in the sample() function
# Note: Accomplishes the same thing as first creating 'n'
random_sample <- fruits[sample(1:length(fruits), size = 5)]
```

<div style="margin-top: 15px"> </div>

Exercises
=============================

Create a vector that represents the age of at least four different family
members or friends. You can name it whatever you want.

1. What is the mean age of the people in your vector? Find out in two ways,
with and without using the `mean()` command.

2. How old is the youngest person in your vector? (Use an R command to find out.)

3. What is the age gap between the youngest person and the oldest person in your vector?
(Again use R to find out, and try to be as general as possible in the sense that
your code should work even if the elements in your vector, or their order, change.)

4. How many people in your vector are above age 25? (Again, try to make your code
work even in the case that your vector changes.)

5. Replace the age of the oldest person in your vector with the age of someone
else you know.

6. Create a new vector that indicates how old each person in your vector
will be in 10 years.

7. Create a new vector that indicates what year each person in your vector
will turn 100 years old.

8. Create a new vector with a random sample of 3 individuals from your
original vector. What is the mean age of the people in this new
vector?

