---
title: Merging &#38; Appending
author: <a href="http://stanford.edu/~ejdemyr/"> Simon Ejdemyr </a> 
date: December, 2015
summary: This tutorial explains how to combine datasets by merging or appending
 them. Merging means transferring *columns* from one dataset to
 another, while appending means transferring *rows* from one dataset to
 another. These are two important tasks since datasets only very rarely
 come with all the information we need. 
output:
  html_document:
     template: ../template/r-tutorial-template-v2.html
     mathjax: null
---



Merging
==================

Say you're interested in how economic development affects democracy
across countries, but all you have are two separate datasets on these
two factors. What you need to do is to combine the two.  When you're
done with this, you also need information on what world region each
country is located in. These tasks can be accomplished with one-to-one
merging and one-to-many merging.


One-to-one merging
--------------------

To illustrate, say you had two datasets that look like this: 

```{r}
# Generate toy dataset 1: 
demo <- data.frame(
    country = c("USA", "Albania", "Turkey", "China", "Sudan"),
    democracy_score = c(19, 16, 16, 5, 10)
    )
demo

# Generate toy dataset 2: 
econ <- data.frame(
    country = c("China", "Albania", "Turkey", "USA", "Sudan"),
    gdp_pc = c(12000, 10000, 9000, 20000, 500)
    )
econ
```

We can merge these datasets using R's `merge()` command, where `by` specifies
the variable the two datasets has in common (usually called an "ID" variable):
```{r}
df <- merge(demo, econ, by = "country")
df
```

You can merge on more than one variable. Say you had two datasets that look 
like this:
 
```{r}
# Generate toy dataset 1: 
demo <- data.frame(
    expand.grid(country = c("USA", "China", "Sudan"), year = 1994:1996),
    democracy_score = round(runif(9, 0, 20), 0)
    )
demo

# Generate toy dataset 2: 
econ <- data.frame(
    expand.grid(year = 1994:1996, country = c("USA", "China", "Sudan")),
    gdp_pc = round(runif(9, 1000, 20000), 0)
    )
econ

merge(demo, econ, by = c("country", "year"))
```

Here, we're merging on both country and year. 

If one of the data frames has missing observations, only observations 
the two data frames have in common will be kept when merging, unless
you specify `all.x`, `all.y`, or `all`: 

```{r}
econ <- econ[-c(5, 9), ] #delete obs. 5 & 9 for illustrative purposes
econ
dim(demo) #different number of observations (rows) than 'econ'
dim(econ) #different number of observations (rows) than 'demo'          

merge(demo, econ, by = c("country", "year")) #keep only matching observations 
merge(demo, econ, by = c("country", "year"), all.x = TRUE) #keep all observations in 'demo'
```


One-to-many merging
------------------------

In the previous examples the ID variables we used for merging uniquely 
identified observations in each data frame. This is called one-to-one
merging. 

Sometimes we need to do things slightly differently, using one-to-many merging. 
Say, for example, we have the following two data frames: 

```{r}
demo <- data.frame(
    expand.grid(country = c("USA", "China", "Sudan"), year = 1994:1996),
    democracy_score = round(runif(9, 0, 20), 0)
    )
demo

region <- data.frame(
    country = c("USA", "China", "Sudan"), 
    region = c("America", "Asia", "Africa")
    )
region
```

A one-to-many merge of these data frames would look like this: 

```{r} 
merge(demo, region, by = "country")
```


Appending
===============

Appending means matching datasets vertically. We can do this in R using 
`rbind()`. The two dataframes you're appending must have identical 
variable names. Here's an example: 

```{r}
df1 <- data.frame(
    year = rep(1990:1995, 2),
    country = c(rep("country1", 6), rep("country2", 6))
    )
df2 <- data.frame(
    year = rep(1996:2000, 2),
    country = c(rep("country1", 5), rep("country2", 5))
    )

df <- rbind(df1, df2) #append

require(dplyr) #to display output in certain order
arrange(df, country, year)
```



Exercises 
==============

**Warm-up:** Merge the three data frames defined below. Your final data
frame should have 9 rows and 5 columns. Hint: merge only two data
frames at a time.

```{r}
df1 <- data.frame(
    name = c("Mary", "Thor", "Sven", "Jane", "Ake", "Stephan",
             "Bjorn", "Oden", "Dennis"),
    treatment_gr = c(rep(c(1, 2, 3), each = 3)),
    weight_p1 = round(runif(9, 100, 200), 0)
    )
df2 <- data.frame(
    name = c("Sven", "Jane", "Ake", "Mary", "Thor", "Stephan", 
             "Oden", "Bjorn"),
    weight_p2 = round(runif(8, 100, 200), 0)
    )                           
df3 <- data.frame(
    treatment_gr = c(1, 2, 3),
    type = c("dog-lovers", "cat-lovers", "all-lovers")
    )

```


**More extensive:** A prominent economic theory predicts that higher
income inequality should be associated with more redistribution from
the rich to the poor (Meltzer and Richard 1981). Let's create a
dataset that will allow us to test this prediction using
U.S. state-level data.

1. Download data on inequality and tax collection (links below). Take
a moment to familiarize yourself with the codebooks. Then read each
dataset into R.
* [Dataset](../data/Frank_All_v97.csv) with six measures of inequality
(originally from [Mark W. Frank](http://www.shsu.edu/eco_mwf/inequality.html)). 
* [Tax data](http://www.census.gov/govs/statetax/historical_data.html) that
can be used to measure fiscal capacity/redistribution. Use 
the "State Government Tax Collections" link. 



2. Keep only the Year, State, Name, Total Taxes, and Total Income
Taxes variables in the tax dataset, and rename these if
necessary. Keep all the variables in the inequality dataset. Subset
both datasets to be in the year range 1960-2012. 

3. Merge the two datasets. Take a moment to think about how to do
this. Your final dataset should have 2650 rows and 12 variables. (Keep
only data on the 50 states; 50 states x 53 years = 2650 observations.)
Hint: You may find [this dataset with state identifiers](../data/State_FIPS.csv) 
helpful. 

4. The Total Tax and Total Income Tax variables from the tax dataset have
commas in them, and are therefore not numeric. Remove the commas and 
convert the variables to numeric. Hint: `gsub()`. 






