---
title: Collapsing Data 
author: <a href="http://stanford.edu/~ejdemyr/"> Simon Ejdemyr </a> 
date: December, 2015
summary: This tutorial explains how to collapse data in R. Collapsing means
 using one or several *grouping variables* to find summary statistics
 --- mean, median, etc. --- for different categories in your data. For
 example, if you have yearly income data for the 50 U.S. states over a
 10-year period (i.e., you have 500 data points), you may want to know
 what the mean income was in each state (collapsing the data to 50 data
 points) or in each year (10 data points). Or you may want to collapse
 the data by year *and* U.S. region, say, South v. non-South (20 data
 points). Like the [tutorial on modifying data](../modifying-data),
 this tutorial draws on a set of intuitive and elegant functions from
 the `dplyr` package.
output:
  html_document:
     template: ../template/r-tutorial-template-v2.html
     mathjax: null
---


Before we begin, let's load the `dplyr` package. We'll make particular
use of two functions from this package: `group_by` and
`summarize`. We'll also make use of chaining, which you can read more
about in the tutorial on
[modifying data](../modifying-data/#chaining).

```{r}
require(dplyr)
```

One grouping variable
==============

To illustrate how collapsing works, let's create a data frame with
three variables: `student` (an id variable that uniquely
identifies each row); `school` (a grouping variable indicating the
student's school); and `sat_score` (the student's SAT score). 

```{r}
grades <- data.frame(
    student = c("al", "bo", "cindy", "dan", "ella", "frank", "gina", "henry"), 
    school = c(rep("stanford", 4), rep("berkley", 4)),
    sat_score = c(750, 730, 690, 800, 780, 720, 730, 700)
    )
```

Which school --- Stanford or Berkley --- attracts students with
a higher SAT score? Based on my (in reality-not-so-random) random
assignment of scores, it appears to be Stanford: 

```{r}
grades %>%
    group_by(school) %>%
    summarize(mean(sat_score))
```

In words, the mean SAT score for Berkley students is 732.5, and the mean
for Stanford students is 742.5. (`dplyr` also outputs some information about
the new data frame for us, such as its dimensions and the class of each
of its variables.)

To be clear, here's how `group_by()` and `summarize()` work. First,
`group_by()` specifies the grouping variable, i.e., the variable that
classifies observations into different clusters. In this case, we're
classifying students by the school they attend. Second, `summarize()`
specifies what to do with the data points within each cluster.  In
this case, we're finding the mean SAT score in each cluster.

It's often useful to assign a name to the collapsed variable:

```{r}
grades %>%
    group_by(school) %>%
    summarize(mean_sat = mean(sat_score))
```

Also note that you'd often want to save the resulting collapsed
data frame to R's memory. Here's how to do this, creating a new
object called `grades_clps` with the collapsed data: 


```{r}
grades_clps <- grades %>%
    group_by(school) %>%
    summarize(mean_sat = mean(sat_score))
grades_clps
```


Several grouping variables
============================

In the previous example we collapsed the data using only one grouping
variable. Collapsing can also be done using several grouping
variables. Let's modify the `grades` data frame to illustrate:

```{r}
grades <- data.frame(
    student = c("al", "bo", "cindy", "dan", "ella", "frank", "gina", "henry"), 
    school = c(rep("stanford", 4), rep("berkley", 4)),
    classof = rep(c(2017, 2017, 2018, 2018), 2), 
    sat_score = c(750, 730, 690, 800, 780, 720, 730, 700)
    )
grades
```

We now have two grouping variables: `school` and `classof`. The latter specifies
the expected graduation year for each student.

Collapsing by these two grouping variables follows the same logic as above. Just
specify the variables to collapse by inside `group_by()`. 

```{r}
grades %>%
    group_by(school, classof) %>%
    summarize(mean_sat = mean(sat_score))
```


Additional manipulation
===========================

One nice thing about using `dplyr` functions for collapsing data is
that you can combine them with other data manipulation functions, many
of which are covered in a separate tutorial on
[modifying data](../modifying-data). The result is elegant code that
is easy to debug and relatively quick to execute. Here's an example in
which I'm filtering the `grades` data frame to class of 2017 and then
collapsing:

```{r}
grades %>%
    filter(classof == 2017) %>%
    group_by(school) %>%
    summarize(mean_sat = mean(sat_score))
```

Here's an example that adds a variable after the collapse (rescaling
the mean SAT scores to be between 0 and 100, assuming 800 is the maximum
possible score):

```{r}
grades %>%
    group_by(school) %>%
    summarize(mean_sat = mean(sat_score)) %>%
    mutate(mean_sat_strd = (mean_sat / 800) * 100)
```


Different functions
====================

In all the examples above I've used `mean()` inside `summarize()`. Of
course you're by no means limited to finding the mean. You can use all of
R's built-in functions or write your own function. Here are examples that
make use of other common functions:

```{r}
grades %>%
    group_by(school) %>%
    summarize(median_sat = median(sat_score),
              sd_sat = sd(sat_score), 
              min_sat = min(sat_score),
              max_sat = max(sat_score),
              dif_maxmin = max_sat - min_sat)
```

Here's an example of using your own function:

```{r}
maxmindif <- function(x) max(x) - min(x)

grades %>%
    group_by(school) %>%
    summarize(dif_maxmin = maxmindif(sat_score))
```

Lastly, `dplyr` provides a few special functions that can be used
within `summarize()`. One very useful special function is `n()`, which
provides the number of observations in each cluster:

```{r}
grades %>%
    group_by(school) %>%
    summarize(no_obs = n())
```


Exercises
=============

1. Read the `world-small.csv` dataset (available
[here](../data/world-small.csv)) into R. Get to know the structure of
this dataset using functions like `dim()`, `head()`, and `summary()`.

2. Find the mean and median GDP per capita and Polity IV score, by
region (that is, for each region in the dataset). Also find the number
of countries by region.

3. Find the mean and median GDP per capita, by region and whether a country
is a "democracy" or not. For the purpose of this exercise, a country is a
"democracy" if it has a Polity IV score of 15 or higher.















